Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-19 doc/leanappendix.tex
Refer to checkers as prototype and verified

\subsection{Evaluation}

@@

We ran the \lean{} checker on the 80 benchmark problems for which the
generated CRAT file had at most 10~million total clauses.  These are the
same benchmarks used in evaluating the optimizations (Section
\ref{app:experiment:optimize}).  Figure~\ref{fig:lean} shows the
comparative performance of the \lean{} checker versus our unverified
checker on the Y axis, with the total number of clauses on the X axis.

As would be expected, the \lean{} checker consistently runs slower
than the unverified checker.  All of the checks completed and confirmed the outcome of the unverified checker.
The ratio of the runtime for \lean{} versus the runtime for
the unverified checker had a harmonic mean of $5.94$.

>>>>

\subsection{Evaluation}
\label{app:lean:evaluation}

@@

We ran the \lean{} checker on the 80 benchmark problems for which the
generated CRAT file had at most 10~million total clauses.  These are the
same benchmarks used in evaluating the optimizations (Section
\ref{app:experiment:optimize}).  Figure~\ref{fig:lean} shows the
comparative performance of the \lean{} checker versus our unverified prototype
checker on the Y axis, with the total number of clauses on the X axis.

As would be expected, the \lean{} checker consistently runs slower
than the prototype checker.  All of the checks completed and confirmed the outcome of the prototype checker.
The ratio of the runtime for \lean{} versus the runtime for
the prototype checker had a harmonic mean of $5.94$.

### 2023-03-19 doc/mc-paper.tex
Refer to checkers as prototype and verified

Our toolchain is not directly
comparable to theirs---they directly certify the output of a model counting
program, while we certify the output of a knowledge compiler and then
produce a certified model count.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

We know of no other
work addressing the question of certifying the output of a knowledge
compiler or model counter.

@@

  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{i=1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\section{Implementations}
We have implemented versions of the three programs that, along with the \dfour{} knowledge compiler,  form the
toolchain illustrated in Figure~\ref{fig:chain}.  One version,
including a proof generator, proof checker, and ring evaluator for
standard and weighted model counting, is designed to be fully
scalable.  The other version, consisting of a proof checker and
ring evaluator has been formally verified within the \lean{} theorem prover.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

@@

Our ring evaluator can perform both standard and weighted model
counting.  It performs arithmetic over a subset of the rationals
we call $\drational$,
consisting of numbers of the form
$a \cdot 2^{b} \cdot 5^{c}$, for integers $a$, $b$, and $c$, and
with $a$ implemented to have arbitrary range.
Allowing scaling by powers of 2 enables the density computation
and rescaling required for standard model counting.  Allowing scaling
by powers of both 2 and 5 enables exact decimal arithmetic, handling
the weights used in the weighted model counting competitions.

@@

We provide only a summary of our experiments here.  A more complete
description is provided in Appendix~\ref{app:experiments}\@. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We then ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graphs generated by \dfour{}.

@@

Running the verified proof checker in \lean{} showed the same scaling
with respect to the proof size as did the unverified checker, but it
required, on average, around 6 times longer.

@@

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
  whether or toolchain can handle challenging benchmark problems,
\item  
the effectiveness of some design choices and optimizations,
\item
the tradeoffs between the unverified and verified checker, and
\item
how the performance of our toolchain compares to that of the MICE model counter verifier.
\end{itemize}

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
our proof generator and unverified checker, limiting proof
generation to 10,000 seconds.  We ran the programs to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

@@

To compare the performance of our toolchain versus that of
MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
generation, proof checking, and ring evaluation in the CRAT toolchain compares to the time to
run their proof generator \progname{nnf2trace} and checker
\progname{sharptrace}.  We add the time for ring evaluation on our
side, since their checker provides a count, although only an
unweighted one.  We ran their tools on the 92 benchmark problems for
which our toolchain requires at most 3000 seconds and retained
the 84 data points for which at least one of the toolchains completes in under 1000 seconds.

>>>>

Our toolchain is not directly comparable to theirs---they directly
certify the numeric value produced by a model counter, while we 
certify that the graph generated by a knowledge compiler is logically
equivalent to the input formula.  We can also validate the unweighted or
weighted count generated from that representation.
%% Their tool
%% chain can be applied to some programs that are pure model counters,
%%while ours can be applied to pure knowledge compilers.
Nonetheless,
we compare the performance of our toolchain to theirs as part of our
experimental evaluation in Appendix~\ref{app:experiments}.

We know of no other
work addressing the question of certifying the output of a knowledge
compiler.

@@

\begin{definition}[Partitioned-Operation Formula]
  A {\em partitioned-operation formula}
 satisfies the following for all product and sum operations:
      \begin{enumerate}
      \item The arguments to each product must have disjoint dependency sets.  That is, operation
        $\bigwedge_{i=1 \leq i \leq k} \phi_i$ requires $\dependencyset(\phi_i) \cap \dependencyset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \item The arguments to each sum must have disjoint models.  That is, operation
        $\bigvee_{i=1 \leq i \leq k} \phi_i$ requires $\modelset(\phi_i) \cap \modelset(\phi_j) = \emptyset$ for $1 \leq i < j \leq k$.
      \end{enumerate}
\end{definition}
     We let $\pand$ and $\por$ denote the product and sum operations in a partitioned-operation formula.

@@

\section{Implementations}
We have implemented versions of the three programs that, along with
the \dfour{} knowledge compiler, form the toolchain illustrated in
Figure~\ref{fig:chain}.  The proof generator is the same in both
cases, since it need not be trusted.
Our {\em verified}
versions of the proof checker and ring evaluator have been formally
verified within the \lean{} theorem prover.  Our long term goal is to
rely on these.  Our {\em prototype} versions are written in C
(checker) and Python (ring evaluator).  These are currently faster and
more scalable, but we anticipate their need will diminish as the
verified version is further optimized.
%% Our long-term goal is to have a toolchain suitable for use in a
%% knowledge compiler competition, along the lines of the annual
%% competitions for other automated reasoning tools.  (The results could be evaluated for correctness
%% either by having them generate checkable proofs or through randomized equivalence tests via function hashing.)
%% Our experiments
%% consider the benchmark problems and computing capabilities that would
%% arise in such a competition and evaluate whether our implementationed toolchain
%% could handle these problems.  As with the original proof-generating
%% SAT competitions, we assume that the proofs at a competition could be
%% checked with the unverified checker.  Having a verified checker
%% that works on some subset of the benchmarks would provide an
%% additional degree of certainty.

@@

Our prototype ring evaluator can perform both standard and weighted
model counting.  It performs arithmetic over a subset of the rationals
we call $\drational$, consisting of numbers of the form $a \cdot 2^{b}
\cdot 5^{c}$, for integers $a$, $b$, and $c$, and with $a$ implemented
to have arbitrary range.  Allowing scaling by powers of 2 enables the
density computation and rescaling required for standard model
counting.  Allowing scaling by powers of both 2 and 5 enables exact
decimal arithmetic, handling the weights used in the weighted model
counting competitions.  For one of the weighted benchmarks, it
generated a result with 260,909 decimal digits.

@@

We provide only a summary of our experiments here.  A more complete
description is provided in Appendix~\ref{app:experiments}\@. For our
evaluation, we used the public benchmark problems from the 2022
standard and weighted model competitions.  We found that there were
180 unique CNF files among these, ranging in size from 250 to
2,753,207 clauses.
%%We ran our programs on a processor with 64~GB of
%%memory and having an attached high-speed, solid-state
%%disk.
With a runtime limit of 4000 seconds, \dfour{} completed for 124 of the
benchmark problems.  Our proof generator was able to convert all of
these into POGs, with their declarations ranging from 304 to
2,761,457,765 (median 848,784) defining clauses.
%(The
%maximum count overflowed the overflowed the 32-bit signed integers we
%used to represent clause identifiers in both the proof generator and
%checker.)
We then ran our proof generator with a time limit of 10,000 seconds.
It was able to generate full proofs for 108 of the problems and
one-sided proofs for an additional 9 of them, leaving just 7 with no
verification.  The prototype checker successfully verified all of the generated
proofs.  The longest runtime for the combination of proof generator
and checker for a full proof was 13,145 seconds.
It is worth noting that, to date, we have not found
any errors in the d-DNNF graphs generated by \dfour{}.

@@

Running the verified proof checker in \lean{} showed the same scaling
with respect to the proof size as did the prototype checker, but it
required, on average, around 5.9 times longer.

@@

Our experiments are designed to evaluate the following:
\begin{itemize}
\item
  whether or toolchain can handle challenging benchmark problems,
\item  
the effectiveness of some design choices and optimizations,
\item
the tradeoffs between the prototype and verified checker, and
\item
how the performance of our toolchain compares to that of the MICE model counter verifier.
\end{itemize}

@@

For each of the 124 d-DNNF graphs generated by \dfour{}, we ran
our proof generator and prototype checker, limiting proof
generation to 10,000 seconds.  We ran the programs to generate and check
one-sided proofs for the graphs, again with a time limit of 10,000
seconds for proof generation.

@@

To compare the performance of our toolchain versus that of
MICE~\cite{fichte:sat:2022}, we consider how the combination of proof
generation, proof checking, and ring evaluation in the CRAT toolchain compares to the time to
run their proof generator \progname{nnf2trace} and checker
\progname{sharptrace}.  We add the time for ring evaluation on our
side, since their checker provides a count, although only an
unweighted one.  We used the prototype versions of the checker and ring evaluator.
We ran their tools on the 92 benchmark problems for
which our toolchain requires at most 3000 seconds and retained
the 84 data points for which at least one of the toolchains completes in under 1000 seconds.
